---
output: github_document
always_allow_html: true
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  fig.retina = 2,
  fig.align = 'center'
)
```

# Rainwater Harvesting and Groundwater Recharge Household Survey - 2025

<!-- badges: start -->

[![License: CC BY
4.0](https://img.shields.io/badge/License-CC_BY_4.0-lightgrey.svg)](https://creativecommons.org/licenses/by/4.0/)

[![DOI](https://zenodo.org/badge/1212554622.svg)](https://doi.org/10.5281/zenodo.19610496)
<!-- badges: end -->

This dataset contains household-level survey data collected in 2025
under the BASEflow monitoring and data systems using the mWater digital
data collection platform. The survey was conducted across selected
communities to assess household water access conditions, perceptions of
groundwater systems, and adoption of rainwater harvesting (RWH)
practices.

The dataset captures both objective household characteristics (such as
location and primary water source) and subjective perceptions and
beliefs regarding groundwater recharge, water scarcity, and the
effectiveness of rainwater harvesting. It also includes behavioural
information on whether households practice rainwater harvesting and the
barriers preventing adoption.

The data is structured at the household level and is suitable for both
quantitative analysis and policy-relevant interpretation.

Key Themes

-   Captured Household water sources and dependency patterns

-   Seasonal water scarcity experiences

-   Community perceptions of groundwater systems

-   Misconceptions and knowledge gaps about groundwater recharge

-   Adoption and non-adoption of rainwater harvesting practices

-   Barriers to rainwater harvesting implementation

-   Perceived solutions and support needs for groundwater recharge

**Potential Use Cases**

1.  Water Resource Planning and Policy

2.  Behaviour Change and Community Education

3.  NGO and Development Programming

4.  Climate Change Adaptation Research

5.  Integration with Hydrogeological Data Systems

## Installation

You can install the development version of rainwaterhsurvey from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("openwashdata/rainwaterhsurvey")
```

```{r}
## Run the following code in console if you don't have the packages
## install.packages(c("dplyr", "knitr", "readr", "stringr", "gt", "kableExtra"))
library(dplyr)
library(knitr)
library(readr)
library(stringr)
library(gt)
library(kableExtra)
```

Alternatively, you can download the individual datasets as a CSV or XLSX
file from the table below.

1.  Click Download CSV. A window opens that displays the CSV in your
    browser.
2.  Right-click anywhere inside the window and select "Save Page As...".
3.  Save the file in a folder of your choice.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

extdata_path <- "https://github.com/openwashdata/rainwaterhsurvey/raw/main/inst/extdata/"

read_csv("data-raw/dictionary.csv") |> 
  distinct(file_name) |> 
  dplyr::mutate(file_name = str_remove(file_name, ".rda")) |> 
  dplyr::rename(dataset = file_name) |> 
  mutate(
    CSV = paste0("[Download CSV](", extdata_path, dataset, ".csv)"),
    XLSX = paste0("[Download XLSX](", extdata_path, dataset, ".xlsx)")
  ) |> 
  knitr::kable()

```

## Data

The package provides access to household-level survey data collected in 2025 under the BASEflow monitoring and data systems using the mWater digital data collection platform. The survey was conducted across selected communities to assess household water access conditions, perceptions of groundwater systems, and adoption of rainwater harvesting (RWH) practices.

```{r}
library(rainwaterhsurvey)
```

### metadata

The dataset `rainwaterhsurvey` has `r nrow(rainwaterhsurvey)`
observations and `r ncol(rainwaterhsurvey)` variables

```{r}
rainwaterhsurvey |> 
  head(3) |> 
  gt::gt() |>
  gt::as_raw_html()
```

For an overview of the variable names, see the following table.

```{r echo=FALSE, message=FALSE, warning=FALSE}
readr::read_csv("data-raw/dictionary.csv") |>
  dplyr::filter(file_name == "metadata.rda") |>
  dplyr::select(variable_name:description) |> 
  knitr::kable() |> 
  kableExtra::kable_styling("striped") |> 
  kableExtra::scroll_box(height = "200px")
```

## Example

```{r}
library(rainwaterhsurvey)

# 1. Knowledge–Practice Gap
# This single visualization shows whether Knowledge translates into action or
# There is a behavioural adoption gap

# Load required libraries
library(dplyr)
library(ggplot2)

# ---- Clean and standardize variables ----
data_clean <- rainwaterhsurvey %>%
  mutate(
    belief_groundwater_replenished = as.logical(belief_groundwater_replenished),
    rwh_practice_binary = ifelse(tolower(rwh_practice) == "yes", TRUE,
                                 ifelse(tolower(rwh_practice) == "no", FALSE, NA))
  )

# ---- Remove NA values separately for each indicator ----
belief_valid <- data_clean %>%
  filter(!is.na(belief_groundwater_replenished))

practice_valid <- data_clean %>%
  filter(!is.na(rwh_practice_binary))

# ---- Calculate percentages (excluding NAs) ----
belief_percent <- mean(belief_valid$belief_groundwater_replenished) * 100
practice_percent <- mean(practice_valid$rwh_practice_binary) * 100

summary_df <- data.frame(
  indicator = c("Belief groundwater replenishable",
                "Practices rainwater harvesting"),
  percentage = c(belief_percent, practice_percent)
)

# ---- Create side-by-side bar chart ----
ggplot(summary_df, aes(x = indicator, y = percentage)) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(aes(label = paste0(round(percentage, 1), "%")),
            vjust = -0.5,
            size = 5) +
  ylim(0, 100) +
  labs(
    title = "Knowledge–Practice Gap",
    subtitle = "Belief in groundwater replenishment vs RWH adoption",
    x = "",
    y = "Percentage of Households"
  ) +
  theme_minimal(base_size = 14)

```

## License

Data are available as
[CC-BY](https://github.com/openwashdata/%7B%7B%7Bpackagename%7D%7D%7D/blob/main/LICENSE.md).

## Citation

Please cite this package using:

```{r}
citation("rainwaterhsurvey")
```
